Predicting Conflict Escalation

### 2023-05-02 docs/hes_final/article.qmd
Formatting Fixes
New Figures and explanations

---
title: "Predicting Conflict Escalation"
author:
  - name: Diana Kolusheva
    email: email@h
    affiliation: Harvard University
  - name: Chad Stoughton
    email: email@h
    affiliation: Harvard University
  - name: Evan Wheeler
    email: email@h
    affiliation: Harvard University
lang: en
format:
  pdf:
    classoption:
    - twocolumn
bibliography: bibliography.bibtex
# https://github.com/quarto-dev/quarto-cli/discussions/2786
include-in-header:
  - text: |
      \usepackage{float}
      \makeatletter
      \let\oldlt\longtable
      \let\endoldlt\endlongtable
      \def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
      \def\longtable@i[#1]{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt[#1]
      }
      \def\longtable@ii{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt
      }
      \def\endlongtable{\endoldlt
      \end{minipage}
      \twocolumn
      \end{figure}}
      \makeatother
# https://tex.stackexchange.com/questions/31301/how-to-reduce-the-number-of-hyphenation
include-before-body:
  - text: |
      \pretolerance=5000
      \tolerance=9000
      \emergencystretch=0pt
      \exhyphenpenalty=100
      \righthyphenmin=4
      \lefthyphenmin=5
embed-resources: true
notebook-links: false
title-block-banner: true
abstract: |
The task of forecasting conflict escalation is of great interest to policy makers, executives, and social scientists. In this article, we propose a novel, text-first approach to computational conflict prediction using contemporary natural language processing and machine learning techniques. We outline three methods to extract generalizable conflict signals directly from a corpus of document embeddings derived from publicly available news sources. The first uses text data enriched by data about the target country, the second uses text data conditioned only on the country label itself, and the third extracts a signal from the text alone, with no additional country or region-specific features. We demonstrate that usable conflict signals can be extracted directly from text sources, and we outline potential avenues for further research that we believe may yield models that can provide early warnings to decision makers to facilitate strategic planning, resource allocation, and disaster mitigation.

@@

### Motivation
Innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children. Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need. Predicting conflict has long been of interest to researchers and practitioners of international relations, and statistical and computational methods to do so can be found in the literature as early as 1944. Today, much of the work in this area has continued to focus on country or regionspecific models to predict conflict.

This project was undertaken in response to the acute needs of the UNICEF Emergency Operations Team. In 2021, UNICEF responded to 483 new and ongoing humanitarian situations in 153 countries (natural disasters, socio-political crises, health emergencies, nutrition crises, etc.) in addition to its global COVID-19 response efforts. This represents a sharp increase since 2018, and the number of disasters continues to climb. With the rising threat of the climate emergency, global shifts toward authoritarianism, and increasing levels of global conflict, innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children. Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need.

@@

@@

>>>>

---
title: "Predicting Conflict Escalation"
author:
  - name: Diana Kolusheva
    email: email@h
    affiliation: Harvard University
  - name: Chad Stoughton
    email: email@h
    affiliation: Harvard University
  - name: Evan Wheeler
    email: email@h
    affiliation: Harvard University
lang: en
format:
  pdf:
    classoption:
    - twocolumn
bibliography: bibliography.bibtex
# https://github.com/quarto-dev/quarto-cli/discussions/2786
include-in-header:
  - text: |
      \usepackage{float}
      \makeatletter
      \let\oldlt\longtable
      \let\endoldlt\endlongtable
      \def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
      \def\longtable@i[#1]{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt[#1]
      }
      \def\longtable@ii{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt
      }
      \def\endlongtable{\endoldlt
      \end{minipage}
      \twocolumn
      \end{figure}}
      \makeatother
# https://tex.stackexchange.com/questions/31301/how-to-reduce-the-number-of-hyphenation
include-before-body:
  - text: |
      \pretolerance=5000
      \tolerance=9000
      \emergencystretch=0pt
      \exhyphenpenalty=100
      \righthyphenmin=4
      \lefthyphenmin=5
embed-resources: true
notebook-links: false
title-block-banner: true
abstract: |
  The task of forecasting conflict escalation is of great interest to policy makers, executives, and social scientists. In this article, we propose a novel, text-first approach to computational conflict prediction using contemporary natural language processing and machine learning techniques. We outline three methods to extract generalizable conflict signals directly from a corpus of ofer 2.5 million document embeddings derived from publicly available news sources. The first uses text data enriched by country context features, the second uses text data conditioned only on the country label itself, and the third extracts a signal from the text alone, with no additional country or region-specific features. We demonstrate that usable conflict signals can be extracted directly from text sources, and we outline potential avenues for further research that we believe may yield models that can provide early warnings to decision makers to facilitate strategic planning, resource allocation, and disaster mitigation.

@@

### Motivation
This project was undertaken in response to the acute needs of the UNICEF Emergency Operations Team. In 2021, UNICEF responded to 483 new and ongoing humanitarian situations in 153 countries (natural disasters, socio-political crises, health emergencies, nutrition crises, etc.) in addition to its global COVID-19 response efforts. This represents a sharp increase since 2018, and the number of disasters continues to climb. With the rising threat of the climate emergency, global shifts toward authoritarianism, and increasing levels of global conflict, innovative approaches are necessary to scale the reach of UNICEF’s impact to meet the increasing needs of vulnerable children. Automated means of identifying and escalating potentially negative situations as they unfold—or before they unfold—can help UNICEF’s emergency response functions be more proactive and efficient in responding quickly to people in need.

@@

Figure 1 shows how this metric captures conflict fluctuations in Ukraine. Prior to the Russian invasion of February 2022, small fluctuations were able to trigger a spike. However, much larger fluctuations in the period following the invasion are filtered out, as the spike formula adjusts to the higher level of baseline conflict. The increase in violence during the Ukrainian counter-offensive in the late summer again triggers a spike, and this reflects significant changes in the situation on the ground.

@@

![Reducing Embedding Dimensionality with Transformer Encoder](imgs/Transformer_Viz.png){width=50%}
